PanLex-RDF

### 2013-01-02 paper/2012/SWJ-PanLex/paper.tex
Some changes at the end

%The same expression may be mentioned in multiple information sources.
%However, an expression's possible meanings depend on the user's interpretion of them in context of such a source.
%For this reason, every meaning depends on an approver.
%\emph{Denotations} capture the mappings between expressions and their possible meanings.  
Note that \emph{approver} combines a user with an \emph{information source},
however the \emph{information source} is not modelled as a distinct entity.
Also, the information of whether two meanings with different approvers
are still equal is not being captured.

@@

\subsection{Transformation workflow}
\label{sec:conversion}
New sources are added to PanLex on almost a daily basis.   
The size of the PanLex database (including indexes) is currently approximately
18 GB, which is big enough to make recurrent RDF conversion cumbersome:
The RDF files of a full conversion of a dataset of this size are very huge.
Using conventional hardware, it takes impractically long
to convert all the data. This makes
testing and debugging and fixing issues in the modelling or data conversion of the
data very time-consuming.
As the PanLex data already resides in a relational database, the use of a virtual
RDB 2 RDF mapping solution is a natural choice. The \emph{Sparqlify Mapping Language}
offers a very easy-to-use mapping language.
Essentially, these mappings consist of three clauses:
\begin{itemize}
  \item \emph{From:} Specifies the logical SQL table (i.e. table, view or query) underlying the Sparqlify-ML view.
  \item \emph{With:} Defines SPARQL variables by means of expressions over relational columns that yield RDF terms.
  \item \emph{Construct:} The template (i.e. the set of triple patterns) to be generated in this view, based on the SPARQL variable definitions.
\end{itemize}
~\autoref{fig:ex:sparqlify-ml} shows an example of a Sparqlify-ML view for the
languages in PanLex: From each row of the \emph{i1} table three resources are
created from the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach also demonstrates that changing the vocubulary or
adding support for new ones does not require an ETL process, and can therefore be done with little effort.
%\todo{Maybe we should link to lexvo directly.}
%PanLex's languages  iso 639-3 code 
%typed with plx:Language
%resources are created for
%the languages in PanLex, linked with \emph{lexvo}, and

@@

\section{Linking}
\label{sec:linking}
In this section we first describe how we interlinked PanLex's language codes
with Lexvo\footnote{\url{http://www.lexvo.org/}} and the expressions with DBpedia\footnote{\url{http://dbpedia.org}}.
Afterwards we give an overview of how the dataset and the related resources are published.

Since Lexvo covers all ISO 639-3 language codes, interlinking could be done with
the Sparqlify-ML view in Figure~\ref{fig:ex:sparqlify-ml}.
For DBpedia we were interested in creating \emph{valid} and
thus \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38} - Note that more languages are available in the raw download section at \url{http://downloads.dbpedia.org/3.8/}},
which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels: 
Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and
removal of punctuation characters were applied.
% PanLex now uses NFC, not NFKD, for character normalization and does not remove punctuation.
The DBpedia resources were then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
Table ~\ref{fig:plx-dbp-link-counts} summarizes the number of links obtained.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{llll}
\toprule
Language & 639-1 & 639-3 & Links \\
\midrule
English    & en & eng & 1,415,241 \\
German     & de & deu &   224,146 \\
French     & fr & fra &   187,364 \\
Italian    & it & ita &   147,485 \\
Spanish    & sp & spa &   117,056 \\
Portuguese & pt & por &   112,266 \\
Polish     & pl & pol &   110,974 \\
Russian    & ru & rus &    68,040 \\
Czech      & cs & ces &    28,767 \\
Catalan    & ca & cat &    27,779 \\
Korean     & ko & kor &    24,912 \\
Turkish    & tr & tur &    22,258 \\
Bulgarian  & bg & bul &    19,431 \\
Hungarian  & hu & hun &    18,203 \\
Slovene    & sl & slv &    11,981 \\
Greek      & el & ell &     1,112 \\
\midrule
Total      &    &     & 2,537,015 \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of DBpedia links per language}
\label{fig:plx-dbp-link-counts}
\end{table}

@@

\section{Publishing}
\label{sec:publishing}
PanLex itself already provides direct access to the PostgreSQL database for scientists,
several user interfaces\footnote{\url{http://panlex.org/try/}} and an
API\footnote{\url{http://panlex.org/tech/doc/api/}}.
%The usage of these interfaces is not restricted in order to make the provided lexical
%translations as accessible as possible.
With our RDF conversion work, we complement these interfaces with
the Linked Data
%\footnote{\url{http://panlex.org/plx/TODO}}
and the SPARQL
endpoint\footnote{\url{http://panlex.org/sparql} and \url{http://panlex.org/snorql}}.
An overview is shown in ~\autoref{fig:panlex-architecture}.
Our code for interlinking is
hosted on GitHub\footnote{\url{https://github.com/AKSW/PanLex-2-RDF}}.
The created linksets are hosted in the PanLex database and published together
with the other data using Sparqlify.

@@

With the RDF version of PanLex, a large lexical resource participates in the Web of Data,
making it available to a broader community. Users can easily explore the PanLex data using
the services provided by SNORQL and Pubby. This also increases the chance of development of
interesting mashups.
For example, the TeraDict translation service\footnote{\url{http://panlex.org/cgi-bin/apidemo-2.cgi}}
could be easily realized using simple SPARQL queries.
% as shown in
%\todo{Create the query: translations in a given target language of an expression}. 
Anothe future usage scenario is to link PanLex to Wortschatz in order to 
improve the recall of word-by-word translations by considering co-occurences.

>>>>

%The same expression may be mentioned in multiple information sources.
%However, an expression's possible meanings depend on the user's interpretion of them in context of such a source.
%For this reason, every meaning depends on an approver.
%\emph{Denotations} capture the mappings between expressions and their possible meanings.  
Note that \emph{approver} combines a user with an \emph{information source},
however the \emph{information source} is not modelled as a distinct entity.
Also, the information of whether or not two meanings with different approvers
are the same is not being captured.

@@

\subsection{Transformation workflow}
\label{sec:conversion}
New sources are added to PanLex on almost a daily basis.   
The size of the PanLex database (including indexes) is currently approximately
18 GB, which is big enough to make recurrent RDF conversion cumbersome:
The RDF file of a full conversion of a database is much larger than the
database itself.
Using conventional hardware, it takes impractically long
to convert all the data. This makes
testing and debugging and fixing issues in the modelling or data conversion of the
data very time-consuming.
As the PanLex data already resides in a relational database, the use of a virtual
RDB 2 RDF mapping solution is a natural choice.
The \emph{Sparqlify system}\footnote{\url{https://github.com/AKSW/Sparqlify}} offers,
besides an efficient query rewriting engine, also a very easy-to-use mapping
language (ML).
Essentially, these mappings consist of three clauses:
\begin{itemize}
  \item \emph{From:} Specifies the logical SQL table (i.e. table, view or query) underlying the Sparqlify-ML view.
  \item \emph{With:} Defines SPARQL variables by means of expressions over relational columns that yield RDF terms.
  \item \emph{Construct:} The template (i.e. the set of triple patterns) to be generated in this view, based on the SPARQL variable definitions.
\end{itemize}
~\autoref{fig:ex:sparqlify-ml} shows an example of a Sparqlify-ML view for the
languages in PanLex: From each row of the \emph{i1} table three resources are
created from the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach also demonstrates that changing the vocubulary or
adding support for new ones does not require an ETL process, and can therefore be done with little effort.
%\todo{Maybe we should link to lexvo directly.}
%PanLex's languages  iso 639-3 code 
%typed with plx:Language
%resources are created for
%the languages in PanLex, linked with \emph{lexvo}, and

@@

\section{Linking}
\label{sec:linking}
The Sparqlify view in the previous section ~\ref{fig:ex:sparqlify-ml} already
established the interlinking of the PanLex languages with Lexvo.
In this section we outline the interlinking with DBpedia.

%In this section we first describe how we interlinked PanLex's language codes
%with Lexvo\footnote{\url{http://www.lexvo.org/}} and the expressions with DBpedia\footnote{\url{http://dbpedia.org}}.
%Afterwards we give an overview of how the dataset and the related resources are published.

%Since Lexvo covers all ISO 639-3 language codes, interlinking could be done with
%the Sparqlify-ML view in Figure~\ref{fig:ex:sparqlify-ml}.
For DBpedia we were interested in creating \emph{valid} and
thus \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38} - Note that more languages are available in the raw download section at \url{http://downloads.dbpedia.org/3.8/}},
which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying
Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and
removal of punctuation characters.
% PanLex now uses NFC, not NFKD, for character normalization and does not remove punctuation.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
Table ~\ref{fig:plx-dbp-link-counts} summarizes the number of links obtained.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{llll}
\toprule
Language & 639-1 & 639-3 & Links \\
\midrule
English    & en & eng & 1,415,241 \\
German     & de & deu &   224,146 \\
French     & fr & fra &   187,364 \\
Italian    & it & ita &   147,485 \\
Spanish    & sp & spa &   117,056 \\
Portuguese & pt & por &   112,266 \\
Polish     & pl & pol &   110,974 \\
Russian    & ru & rus &    68,040 \\
Czech      & cs & ces &    28,767 \\
Catalan    & ca & cat &    27,779 \\
Korean     & ko & kor &    24,912 \\
Turkish    & tr & tur &    22,258 \\
Bulgarian  & bg & bul &    19,431 \\
Hungarian  & hu & hun &    18,203 \\
Slovene    & sl & slv &    11,981 \\
Greek      & el & ell &     1,112 \\
\midrule
Total      &    &     & 2,537,015 \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of DBpedia links per language}
\label{fig:plx-dbp-link-counts}
\end{table}

@@

\section{Publishing}
\label{sec:publishing}
PanLex itself already provides direct access to the PostgreSQL database for scientists,
several user interfaces\footnote{\url{http://panlex.org/try/}} and an
API\footnote{\url{http://panlex.org/tech/doc/api/}}.
%The usage of these interfaces is not restricted in order to make the provided lexical
%translations as accessible as possible.
With our RDF conversion work, we complement these interfaces with
the Linked Data
%\footnote{\url{http://panlex.org/plx/TODO}}
and the SPARQL
endpoint\footnote{\url{http://panlex.org/sparql} and \url{http://panlex.org/snorql}}.
An overview is shown in ~\autoref{fig:panlex-architecture}.
Our Sparqlify-ML views and the code for interlinking are
hosted on GitHub\footnote{\url{https://github.com/AKSW/PanLex-2-RDF}}.
The created linksets are hosted in the PanLex database and published together
with the other data using Sparqlify.

@@

With the RDF version of PanLex, a large lexical resource participates in the Web of Data,
making it available to a broader community. Users can easily explore the PanLex data using
the services provided by SNORQL and Pubby. This also increases the chance of development of
new interesting mashups.
For example, the TeraDict translation service\footnote{\url{http://panlex.org/cgi-bin/apidemo-2.cgi}}
could now be easily realized using simple SPARQL queries.
% as shown in
%\todo{Create the query: translations in a given target language of an expression}. 
Another future usage scenario is to link PanLex to Wortschatz with the aim of 
improving the recall of word-by-word translations by considering the co-occurences.
